The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-28 article_final/main.tex
made some corrections considering language and notation

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^k) \in \argmin_{\mD \in \cS} \<\mM^k, \mD>\,,
\end{equation}
where $\mM^k$ is a gradient (or a momentum) of $F$ in $\mX^k$ and $\cS \subset \Rmn$ is some set. The update of the algorithm is defined as follows:
\begin{equation}\label{eq:simple_update}
	\mX^{k + 1} = \mX^k + \gamma_k\mathrm{LMO}\left(\mM^k\right)\,.
\end{equation}

We are particularly interested in the case when $\cS$ is a unit ball in the norm $\norm{\cdot}$:

\[
\cS = \cB_{\norm{\cdot}} = \{ \mD \in \Rmn \mid \norm{\mD} \leq 1 \}\,.
\]
In this case,
\[
\argmin_{\mD \in \cS} \<\mM^k, \mD> = - \{ \mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,,
\]
and update for $\mX^k$ in \equationref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}

Using this formula it is easy to compute updates for algorithms induced by various norms $\norm{\cdot}$:

\paragraph{Frobenius norm and Normalized SGD}
When the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \equationref{eq:our_update} turns into
\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \frac{\mM^k}{\normf{\mM^k}}\,,
\end{equation}
which recovers Normalized SGD (NSGD).

\paragraph{Spectral norm and Muon}
When the norm is the spectral norm $\norms{\cdot}$, we get
\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta_k \mU \mV^\top\,,
\end{equation}
which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [\vu_1, \vu_2, \dots, \vu_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and $\mV = [\vv_1, \vv_2, \dots, \vv_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.

\paragraph{Chebyshev norm and SignSGD}
When the norm is the Chebyshev norm $\normc{\cdot}$, we get
\begin{equation}\label{eq:signsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \sign(\mM^k)\,,
\end{equation}
which recovers SignSGD~\citep{bernstein2018signsgd}. Here, $\sign(\mM^k)$ denotes the element-wise sign function. SignSGD is particularly notable for its communication efficiency in distributed training, as it compresses gradients to 1-bit per parameter.

@@

\subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
After considering $\normf{\mM^k}$ and $\norms{\mM^k}$, it is natural to look at the nuclear norm $\normn{\mM^k}$.
\begin{lemma}\label{lemma:neon_update}
	When $\norm{\cdot} = \normn{\cdot}$, \equationref{eq:our_update} becomes
	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta_k \vu_1 \vv_1^\top\,.
	\end{equation}
\end{lemma}
\begin{proof}
Since $\norm{\cdot}^\dagger = \norms{\cdot}$, the goal is to reach $\norms{\mM^k} = \sigma_1$. Let us note that $\Delta = \vu_1 \vv_1^\top$ delivers this value. Indeed, $\normn{\Delta}=1$ and by the trace property and orthogonality of the singular vectors,
\[
\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \vu_1 \vv_1^\top> = \tr \diag({\sigma_1, 0, \dots, 0}) = \norms{\mM^k}\,,
\]
which completes the proof.
\end{proof}

We name the derived algorithm \emph{Neon}. In \sectionref{sec:matrix-side}, we will discuss how to compute the Neon's update.

@@

However, Schatten norms do not fill the gap rank: indeed, when $p > 1$, the rank of the update is full, while when $p=1$ ($p \rightarrow 1$) it is one. Moreover, it is not clear how to calculate the update for $p \neq 1, 2, \infty$: it seems one has to know all $\sigma_i$ to compute the update, which makes the problem as hard as the full SVD.

\subsubsection{Ky Fan norms}
There is another family of matrix norms, which might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e. the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the spectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

Let us note that $\Delta = \sum_{i=1}^{k} \vu_i \vv_i^\top$ delivers the value. Indeed,
\[
\<\mM^t, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} \vu_i \vv_i^\top> = \sum_{i,j=1}^{r, k}\<\vu_i \sigma_i \vv_i^\top, \vu_j \vv_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}\,,
\]
which completes the proof.
\end{proof}
These updates constitute the family of \emph{Fanions}, LMO-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. The algorithm for a particular $k$ we will call \emph{Fanion-$k$}. Thus, Muon is a Fanion-$\min\{m,n\}$, while Neon is a Fanion-1. Moreover, the unsharded version of the rank-$r$ Dion (Algorithm 1 from \citet{ahn2025dion}) without the error feedback and without scaling of the update is actually a Fanion-$r$ (see \citet{pethick2025understandingdion}, where Dion is written down in a notation more similar to ours).

@@

Simply applying norm dual to the Ky Fan $k$-norm however, is not enough to provide family of algorithms diverse enough for our cause. Next we consider linear combinations of algorithms from LMO framework, which happen to be LMO-algorithms too, as we show in succeeding paragraphs.

>>>>

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot} = \<\cdot,\cdot>$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^k) \in \argmin_{\mD \in \cS} \<\mM^k, \mD>\,,
\end{equation}
where $\mM^k$ is the gradient (or momentum) of $F$ at $\mX^k$ and $\cS \subset \Rmn$ is some set. The update of the algorithm is defined as follows:
\begin{equation}\label{eq:simple_update}
	\mX^{k + 1} = \mX^k + \gamma_k\mathrm{LMO}\left(\mM^k\right)\,.
\end{equation}

We are particularly interested in the case when $\cS$ is a unit ball in some norm $\norm{\cdot}$:
\[
\cS = \cB_{\norm{\cdot}} = \{ \mD \in \Rmn \mid \norm{\mD} \leq 1 \}\,.
\]
In this case,
\[
\argmin_{\mD \in \cS} \<\mM^k, \mD> = - \{ \mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,,
\]
and the update for $\mX^k$ in \equationref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}

Using this formula, it is easy to compute updates for algorithms induced by various norms $\norm{\cdot}$:

\paragraph{Frobenius norm and Normalized SGD}
When the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \equationref{eq:our_update} turns into
\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \frac{\mM^k}{\normf{\mM^k}}\,,
\end{equation}
which recovers Normalized SGD (NSGD).

\paragraph{Spectral norm and Muon}
When the norm is the spectral norm $\norms{\cdot}$, we get
\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \mU \mV^\top\,,
\end{equation}
which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [\vu_1, \vu_2, \dots, \vu_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and $\mV = [\vv_1, \vv_2, \dots, \vv_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.

\paragraph{Chebyshev norm and SignSGD}
When the norm is the Chebyshev norm $\normc{\cdot}$, we get
\begin{equation}\label{eq:signsgd_update}
	\mX^{k+1} = \mX^{k} - \gamma_k \sign(\mM^k)\,,
\end{equation}
which recovers SignSGD~\citep{bernstein2018signsgd}. Here, $\sign(\mM^k)$ denotes the element-wise sign function. SignSGD is particularly notable for its communication efficiency in distributed training, as it compresses gradients to 1-bit per parameter.

@@

\subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
After considering $\normf{\mM^k}$ and $\norms{\mM^k}$, it is natural to look at the nuclear norm $\normn{\mM^k}$.
\begin{lemma}\label{lemma:neon_update}
	When $\norm{\cdot} = \normn{\cdot}$, \equationref{eq:our_update} becomes
	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \gamma_k \vu_1 \vv_1^\top\,.
	\end{equation}
\end{lemma}
\begin{proof}
Since $\norm{\cdot}^\dagger = \norms{\cdot}$, the goal is to reach $\norms{\mM^k} = \sigma_1$. Let us note that $\Delta = \vu_1 \vv_1^\top$ delivers this value. Indeed, $\normn{\Delta}=1$ and by the trace property and orthogonality of the singular vectors,
\[
\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \vu_1 \vv_1^\top> = \tr \diag({\sigma_1, 0, \dots, 0}) = \norms{\mM^k}\,,
\]
which completes the proof.
\end{proof}

We name the derived algorithm \emph{Neon}. In \sectionref{sec:matrix-side}, we will discuss how to compute Neon's update.

@@

However, Schatten norms do not fill the rank gap: indeed, when $p > 1$, the rank of the update is full, while when $p=1$ ($p \rightarrow 1$) it is one. Moreover, it is not clear how to calculate the update for $p \neq 1, 2, \infty$: it seems one has to know all $\sigma_i$ to compute the update, which makes the problem as hard as computing the full SVD.

\subsubsection{Ky Fan norms}
There is another family of matrix norms that might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e., the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the spectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

Let us note that $\Delta = \sum_{i=1}^{k} \vu_i \vv_i^\top$ delivers the value. Indeed,
\[
\<\mM^t, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} \vu_i \vv_i^\top> = \sum_{i,j=1}^{r, k}\<\vu_i \sigma_i \vv_i^\top, \vu_j \vv_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}\,,
\]
which completes the proof.
\end{proof}
These updates constitute the family of \emph{Fanions}, LMO-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. We call the algorithm for a particular $k$ \emph{Fanion-$k$}. Thus, Muon is Fanion-$\min\{m,n\}$, while Neon is Fanion-1. Moreover, the unsharded version of rank-$r$ Dion (Algorithm 1 from \citet{ahn2025dion}) without error feedback and without scaling of the update is actually Fanion-$r$ (see \citet{pethick2025understandingdion}, where Dion is written in a notation more similar to ours).

@@

Simply applying the norm dual to the Ky Fan $k$-norm, however, is not enough to provide a family of algorithms diverse enough for our purposes. Next, we consider linear combinations of algorithms from the LMO framework, which happen to be LMO-algorithms as well, as we show in the following paragraphs.
